On the instability and degeneracy of deep learning models

### 2019-06-11 writing/paper-note/docs/note.tex
started edits

@@

\begin{abstract}
{A probability model exhibits instability if small changes in a data
outcome result in large, and often unanticipated, changes in
probability. This instability is a property of the probability model,
rather than the fitted parameter vector. For correlated data structures
found in several application areas, there is increasing interest in
predicting/identifying such sensitivity in model probability structure.
We consider the problem of quantifying instability for general
probability models defined on sequences of observations, where each
sequence of length \(N\) has a finite number of possible values. A
sequence of probability models results, indexed by \(N\), that
accommodates data of expanding dimension. Model instability is formally
shown to occur when a certain log-probability ratio under such models
grows faster than \(N\). In this case, a one component change in the
data sequence can shift probability by orders of magnitude. Also, as
instability becomes more extreme, the resulting probability models are
shown to tend to degeneracy, placing all their probability on
potentially small portions of the sample space. These results on
instability apply to large classes of models commonly used in random
graphs, network analysis, and machine learning contexts.}
{Degeneracy, Instability, Classification, Deep Learning, Graphical Models}
\end{abstract}
% end imaiai stuff

@@

We consider the behavior, and the potential impropriety, of sequences of
discrete probability models built to incorporate observations of
increasing sample size \(N\). Interest is in identifying instability in
such models, which is roughly characterized by probabilities with
extreme sensitivity to small changes in data configuration. The concept
of instability was introduced in the field of statistical physics (i.e.,
point processes) by \citet{ruelle1999statistical} and then further
extended by \citet{schweinberger2011instability} for a family of
exponential models. At issue, models exhibiting instability are
typically undesirable as these tend to provide poor representations of
data or data-generation. As an example, such models can include
near-degenerate distributions that assign essentially all probability
mass to only a subset of an overall sample space. The latter issue in
connection to degeneracy has been recognized as a concern in that
dominant model outcomes may not resemble observed data
\citep[cf.][]{handcock2003assessing}. As a compounding issue, model
instability often has direct negative impacts for statistical inference
and computations based on likelihood functions. Namely, volatilities in
probability structure can potentially hamper the numerical evaluations
required for maximum likelihood estimation as well as other model-based
simulations via Markov Chain Monte Carlo (MCMC). These reasons motivate
our general study of instability for a broad class of probability
models, described next.

In the model framework, let \(\boldsymbol X_N = (X_1, \dots, X_N)\)
denote a collection of discrete random variables with a finite sample
space, \(\mathcal{X}^N\), represented as some \(N\)-fold Cartesian
product. That is, \(\mathcal{X}\) with \(|\mathcal{X}| < \infty\)
denotes the set of potential outcomes for each single variable \(X_i\),
so that the product space \(\mathcal{X}^N\) corresponds to values for
the variables \(\boldsymbol X_N=(X_1,\ldots,X_N)\). For each \(N\), let
\(P_{\boldsymbol \theta_N}\) denote a probability model on
\(\mathcal{X}^N\), under which
\(P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0\) is the probability of
the data outcome \((x_1, \dots, x_N) \in \mathcal{X}^N\). In this, we
assume that the model support of \(P_{\boldsymbol \theta_N}\) is the
sample space \(\mathcal{X}^N\). This framework produces probability
models \(P_{\boldsymbol \theta_N}\), indexed by a generic sequence of
parameters \(\boldsymbol \theta_N\), to describe data
\(\boldsymbol X_N\) of any given sample size \(N \geq 1\). For
simplicity, we will refer to this distributional class as \emph{Finite
Outcome Everywhere Supported (FOES)} models in the following. The
dimension and structure of such parameters are generic, without
restriction, though natural cases will be seen to include those where
\(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) for some arbitrary
integer-valued function \(q(\cdot) \geq 1\).

@@

For random variables
\(\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)\) with sample
space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an
exponential family model for \(\boldsymbol X\) with probability mass
function given by \begin{equation}
\label{eq:expo}
p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\end{equation} depending on parameter vector
\(\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}\) and natural
parameter function
\(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed
positive integers \(k\) and \(L\) denoting their dimensions. Above,
\(\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L\) is a vector of
sufficient statistics, while \[
\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
\] denotes the normalizing function with parameter space \(\Theta_N\).
The natural parameter function \(\eta (\boldsymbol \theta)\) has a
linear form (i.e.,
\(\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta\) for a given
\(L \times k\) matrix \(\bm{A}\)) in many common model formulations,
though may also be nonlinear (e.g., curved exponential families). In the
linear case, \(\eta (\boldsymbol \theta) = \boldsymbol \theta\) may be
generally assumed in the exponential parameterization with a minor
modification to the definition of sufficient statistics
\(\boldsymbol g_N(\boldsymbol x)\).

Such discrete exponential family models are special cases of the FOES
models, as seen by defining
\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0\),
\(\boldsymbol x \in \mathcal{X}^N\), based on \eqref{eq:expo} and a
parameter sequence
\(\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k\). For example,
if observations \(\boldsymbol X = (X_1,\ldots,X_N)\) correspond to \(N\)
independent and identically distributed Bernoulli random variables, each
indicating a binary \(0\)-\(1\) outcome, the resulting probabilities
have exponential form \eqref{eq:expo} given by \begin{equation}

@@

 \end{equation} with sufficient statistic
\(\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N x_i\) and ``log odds
ratio'' parameter
\(\boldsymbol \theta_N \equiv \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}\).
More generally, supposing \(\boldsymbol X =(X_1,\ldots,X_N)\) represent
\(N\) independent trials, each assuming an outcome \(\{1,\ldots,k\}\)
among \(k\) possibilities (e.g., a die roll), a multinomial distribution
is given by \begin{equation}

@@

\end{equation} with sufficient statistic
\(\boldsymbol g_N(\boldsymbol x)\) involving a count
\(\sum_{i=1}^N \mathbb{I}(x_i=j)\) for each outcome
\(j \in \{1,\ldots,k\}\), where \(\mathbb{I}(\cdot)\) denotes the
indicator function, and parameters
\(\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k\)
defining log-probability ratios
\(\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]\).
In addition to such standard models for discrete independent data,
exponential models of FOES type commonly arise with dependent spatial
data \citep{besag1974spatial} and network/relational data
\citep{wasserman1994social, handcock2003assessing}. For a random graph
or network with, say, \(n\) nodes, consider \(N={n \choose 2}\) random
edges where the \(i\)th edge is associated with a pair of nodes
\(s_i \equiv \{v_i,u_i\}\) and a binary variable \(X_i\in\{0,1\}\)
indicating presence/absence of an edge among the node pair \(s_i\),
\(i=1,\ldots,N\). Here the length \(N\) of the edge variable sequence
\(\boldsymbol X = (X_1,\ldots,X_N)\) increases as a function of node
number \(n\) and corresponding exponential models often incorporate
graph topographical features derived from \(\boldsymbol X\). As an
example, consider a graph model of exponential/FOES form prescribed by

@@

\end{equation} \[

@@

\] involving the numbers of edges, 2-stars and triangles among an
outcome \(\boldsymbol x\) given by \(g_{1,N}(\boldsymbol x)\),
\(g_{2,N}(\boldsymbol x)\) and \(g_{3,N}(\boldsymbol x)\), respectively,
along with \(k=3\) real parameters
\(\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})\).
For this network model \eqref{eq:mod2} in particular, as well as for more
general models of form \eqref{eq:expo},
\citet{schweinberger2011instability} considered instability in such
exponential models with sequences of fixed parameters
\(\boldsymbol \theta_N = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k\),
\(N \geq 1\), of fixed dimension \(k\).

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model
specified for discrete or continuous random variables, with binary
variables being most common \citep[cf.][]{smolensky1986information}. A
RBM architecture has two layers, hidden (\(\mathcal{H}\)) and visible
(\(\mathcal{V}\)), with conditional independence within each layer. Let
\(\boldsymbol X = (X_1,\ldots,X_N)\) denote the \(N\) random variables
for visibles with support \(\mathcal{X}^N\) and
\(\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})\) denote the
\(N_\mathcal{H}\) random variables for hiddens with support
\(\mathcal{X}^{N_\mathcal{H}}\) where \(\mathcal{X} = \{-1,1\}\). For
parameters
\(\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}\),
\(\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N\), and
\(\boldsymbol \theta_N^{\mathcal{HV}}\) as a real matrix with dimension
\(N_\mathcal{H} \times N\), the RBM model for
\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) has the joint
probability mass function \begin{equation}

@@

\end{equation} with normalizing function \[

@@

\] Let
\(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}\),
with \(q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}\), denote the
parameter vector for the RBM, as indexed by the number \(N\) of visible
random variables (which may differ from the actual lengths of these
parameter vectors). The probability mass function for the visible
variables \(\boldsymbol X = (X_1, \dots, X_N)\) follows from
marginalizing the joint specification to yield \begin{equation}

@@

\end{equation} Here the baseline model \eqref{eq:RBM1} for hidden/visible
variables is a linear exponential one in sufficient statistics
\((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) using
\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) from
\eqref{eq:RBM1}, but the form differs from the previous exponential models
in \eqref{eq:expo} in that the lengths of parameters
\(\boldsymbol \theta_N\) and statistics
\((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) increase to
incorporate more visible variables. That is, in contrast to
\eqref{eq:expo}, the natural parameter function involved in the RBM model
\eqref{eq:RBM1}, as the identity mapping of the parameters
\(\boldsymbol \theta_N\in\mathbb{R}^{q(N)}\), naturally grows in
dimension \(q(N)\to \infty\) to accommodate visible variables
\(X_1, \dots, X_N\) of increasing sample size \(N\to \infty\).
Additionally, one may further arbitrarily choose the number
\(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RBM
model \eqref{eq:RBM1} to define a marginal model \eqref{eq:RBM2} for the
\(N\) visible variables \(\boldsymbol X\), and the number
\(N_\mathcal{H}\) of hiddens may also potentially increase with \(N\).
Because \(|\mathcal{X}| = 2\) and
\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all
\(\boldsymbol x \in \mathcal{X}^N\), the RBM specification \eqref{eq:RBM2}
for visibles \(\boldsymbol\) corresponds to a FOES model, while the
joint distribution \eqref{eq:RBM1} for \((\boldsymbol X, \boldsymbol H)\)
is also a FOES model. As this example also indicates, any model formed
by marginalizing a base FOES model class, such as the RBM joint
specification \eqref{eq:RBM1}, is again a FOES model.

@@

Consider two models with ``deep architecture'' that contain multiple
hidden (or latent) layers in addition to a visible layer of data, namely
a deep Boltzmann machine \citep{salakhutdinov2009deep} and a deep belief
network \citet{hinton2006fast}{]}. Let \(M\) denote the number of hidden
layers included in the model and let \(N_{(H,1)}, \dots, N_{(H,M)}\)
denote the numbers of hidden variables within each hidden layer. Then
the random vector
\(\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}\)
collects both the hidden variables
\(\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}\) and visible
variables \(\boldsymbol X =(X_1,\ldots,X_N)\) in a deep probabilistic
model. Each variable outcome will again lie in
\(\mathcal{X} = \{-1,1\}\).

@@

\textbf{Deep belief network (DBN).} A DBN resembles a DBM in that there
are multiple layers of latent random variables stacked in a deep
architecture with no conditional dependence between layers. The
difference between the DBM and DBN models is that all but the last
stacked layer in a DBN are Bayesian networks
\citep[see][]{pearl985bayesian}, rather than RBMs. A Bayesian network is
a probabilistic graphical model that defines conditional dependence to
be directed, rather than undirected (as with the RBM). Thus for visibles
\(X_1, \dots, X_N\) with support
\(\mathcal{X}^N, \mid \mathcal{X} \mid < \infty\), a DBN is also a FOES
model with \(q(N)\) the length of parameter vector is dependent on the
dimension of the visibles because
\(P_{\boldsymbol \theta_N}(\boldsymbol x)>0\) for all
\(\boldsymbol x \in\mathcal{X}^N\). Commonly, as in logistic belief nets
\citep{neal1992connectionist}, a ``weight'' parameter is placed on each
interaction between visibles, \(X_1, \dots, X_N\), and the first layer
of latent variables, \(H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}\),
satisfying the definition of a FOES model.

@@

To define a measure of instability in FOES models, it is useful to
consider the behavior of data models \(P_{\theta_N}\), again supported
on a set \(\mathcal{X}^N\) of outcomes for
\(\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)\), in connection
to the sample size \(N\). A relevant quantity to this end is a log-ratio
of extremal probabilities (LREP), defined as \begin{align}
\label{eq:elpr}
 \REP (\boldsymbol \theta_N)  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right],
\end{align} based on maximum and minimal model probabilities. In what
follows, the main idea is that instability, and other negative model
features, can be associated with a FOES model formulation for \(N\)
random variables where the LREP \eqref{eq:elpr} is overly large relative
to the sample size \(N\). That is, a sequence of FOES probability models
\(P_{\boldsymbol \theta_N}\) results in specifying the distribution of
observations \(\boldsymbol X=(X_1,\ldots,X_N)\) for each sample size
\(N \geq 1\) and instability will generally occur among these models
whenever the corresponding LREP \eqref{eq:elpr} grows faster than \(N\).
This leads to the following definition.

\BeginKnitrBlock{definition}[S-unstable FOES model]
\protect\hypertarget{def:instabFSFS}{}{\label{def:instabFSFS}
\iffalse (S-unstable FOES model) \fi{} }A FOES model formulation for
\(\boldsymbol X_N=(X_1,\ldots,X_N)\) is \emph{Schweinberger-unstable} or
\emph{S-unstable} if \begin{equation}
\label{eq:Sun}
\lim \limits_{N \rightarrow \infty} \frac{1}{N} \REP(\boldsymbol \theta_N) \equiv \lim \limits_{N \rightarrow \infty} \frac{1}{N}\log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right] = \infty
\end{equation} as the number of variables increases
(\(N \rightarrow \infty\)).
\EndKnitrBlock{definition}

@@

\] and \begin{align*}

@@

\end{align} where

>>>>

\usepackage{color}
\newcommand{\ak}[1]{{\color{blue} #1}}

@@

\begin{abstract}
{A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. This instability is a property of the probability \ak{model, prescribed by a probability distributional form evaluated at a given configuration of parameters}. For correlated data structures found in several application areas, there is increasing interest in \ak{identifying} such sensitivity in model probability structure. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length \(N\) has a finite number of possible values \ak{that can be taken at each point}. A sequence of probability models results, indexed by \(N\), \ak{and an associated parameter sequence,} that accommodates data of expanding dimension. Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than \(N\). In this case, a one component change in the data sequence can shift probability by orders of magnitude. Also, as instability becomes more extreme, the resulting probability models are shown to tend to degeneracy, placing all their probability on potentially small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.}
{Degeneracy, Instability, Classification, Deep Learning, Graphical Models}
\end{abstract}
% end imaiai stuff

@@

We consider the behavior, and the potential impropriety, of sequences of discrete probability models built to incorporate observations of increasing sample size \(N\). Interest is in identifying instability in such models, which is roughly characterized by probabilities with extreme sensitivity to small changes in data configuration. The concept of instability was introduced in the field of statistical physics (i.e., point processes) by \citet{ruelle1999statistical} and then further extended by \citet{schweinberger2011instability} for a family of exponential models. At issue, models exhibiting instability are typically undesirable as these tend to provide poor representations of data or data-generation. As an example, such models can include near-degenerate distributions that assign essentially all probability mass to only a subset of an overall sample space. The latter issue in connection to degeneracy has been recognized as a concern in that dominant model outcomes may not resemble observed data \citep[cf.~][]{handcock2003assessing}. As a compounding issue, model instability often has direct negative impacts for statistical inference and computations based on likelihood functions. Namely, volatilities in probability structure can potentially hamper the numerical evaluations required for maximum likelihood estimation as well as other model-based simulations via Markov Chain Monte Carlo (MCMC). These reasons motivate our general study of instability for a broad class of probability models, described next.

In the model framework, let \(\boldsymbol X_N = (X_1, \dots, X_N)\) denote a collection of discrete random variables with a finite sample space, \(\mathcal{X}^N\), represented as some \(N\)-fold Cartesian product. That is, \(\mathcal{X}\) with \(|\mathcal{X}| < \infty\) denotes the set of potential outcomes for each single variable \(X_i\), so that the product space \(\mathcal{X}^N\) corresponds to values for the variables \(\boldsymbol X_N=(X_1,\ldots,X_N)\). For each \(N\), let \(P_{\boldsymbol \theta_N}\) denote a probability model on \(\mathcal{X}^N\), under which \(P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0\) is the probability of the data outcome \((x_1, \dots, x_N) \in \mathcal{X}^N\). In this, we assume that the model support of \(P_{\boldsymbol \theta_N}\) is the sample space \(\mathcal{X}^N\). This framework produces probability models \(P_{\boldsymbol \theta_N}\), indexed by a \ak{defining} sequence of parameters \(\boldsymbol \theta_N\), to describe data \(\boldsymbol X_N\) of any given sample size \(N \geq 1\). For simplicity, we will refer to this distributional class as \emph{Finite Outcome Everywhere Supported (FOES)} models in the following. The dimension and structure of such parameters are generic, without restriction, though natural cases will be seen to include those where \(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) for some arbitrary integer-valued function \(q(\cdot) \geq 1\).

@@

For random variables \(\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)\) with sample space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an exponential \ak{model family} for \(\boldsymbol X\) with probability mass function given by
\begin{equation}
\label{eq:expo}
p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\end{equation}
depending on parameter vector \(\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}\) and natural parameter function \(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed positive integers \(k\) and \(L\) denoting their dimensions. Above, \(\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L\) is a vector of sufficient statistics, while
\[
\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
\]
denotes the normalizing function with parameter space \(\Theta_N\). The natural parameter function \(\eta (\boldsymbol \theta)\) has a linear form (i.e., \(\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta\) for a given \(L \times k\) matrix \(\bm{A}\)) in many common model formulations, though may also be nonlinear (e.g., curved exponential families). In the linear case, \(\eta (\boldsymbol \theta) = \boldsymbol \theta\) may be generally assumed in the exponential parameterization with a minor modification to the definition of sufficient statistics \(\boldsymbol g_N(\boldsymbol x)\).

Such discrete \ak{exponential models} are special cases of the FOES models, as seen by defining \(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0\), \(\boldsymbol x \in \mathcal{X}^N\), based on \eqref{eq:expo} and a parameter sequence \(\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k\). For example, if observations \(\boldsymbol X = (X_1,\ldots,X_N)\) correspond to \(N\) independent and identically distributed Bernoulli random variables, each indicating a binary \(0\)-\(1\) outcome, the resulting probabilities have exponential form \eqref{eq:expo} given by

@@

 \end{equation}
with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N x_i\) and ``log odds ratio'' parameter \(\boldsymbol \theta_N \equiv \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}\). More generally, supposing \(\boldsymbol X =(X_1,\ldots,X_N)\) represent \(N\) independent trials, each assuming an outcome \(\{1,\ldots,k\}\) among \(k\) possibilities (e.g., a die roll), a multinomial distribution is given by

@@

\end{equation}
with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\) involving a count \(\sum_{i=1}^N \mathbb{I}(x_i=j)\) for each outcome \(j \in \{1,\ldots,k\}\), where \(\mathbb{I}(\cdot)\) denotes the indicator function, and parameters \(\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k\) defining log-probability ratios \(\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]\). In addition to such standard models for discrete independent data, exponential models of FOES type commonly arise with dependent spatial data \citep{besag1974spatial} and network/relational data \citep{wasserman1994social, handcock2003assessing}. For a random graph or network with, say, \(n\) nodes, consider \(N={n \choose 2}\) random edges where the \(i\)th edge is associated with a pair of nodes \(s_i \equiv \{v_i,u_i\}\) and a binary variable \(X_i\in\{0,1\}\) indicating presence/absence of an edge among the node pair \(s_i\), \(i=1,\ldots,N\). Here the length \(N\) of the edge variable sequence \(\boldsymbol X = (X_1,\ldots,X_N)\) increases as a function of node number \(n\) and corresponding exponential models often incorporate graph topographical features derived from \(\boldsymbol X\). As an example, consider a graph model of exponential/FOES form prescribed by

@@

\end{equation}
\[

@@

\]
involving the numbers of edges, 2-stars and triangles among an outcome \(\boldsymbol x\) given by \(g_{1,N}(\boldsymbol x)\), \(g_{2,N}(\boldsymbol x)\) and \(g_{3,N}(\boldsymbol x)\), respectively, along with \(k=3\) real parameters \(\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})\). For this network model \eqref{eq:mod2} in particular, as well as for more general models of form \eqref{eq:expo}, \citet{schweinberger2011instability} considered instability in such exponential models with sequences of fixed parameters \(\boldsymbol \theta_N = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k\), \(N \geq 1\), of fixed dimension \(k\).

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, with binary variables being most common \citep[cf.~][]{smolensky1986information}. A RBM architecture has two layers, hidden (\(\mathcal{H}\)) and visible (\(\mathcal{V}\)), with conditional independence within each layer. Let \(\boldsymbol X = (X_1,\ldots,X_N)\) denote the \(N\) random variables for visibles with support \(\mathcal{X}^N\) and \(\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})\) denote the \(N_\mathcal{H}\) random variables for hiddens with support \(\mathcal{X}^{N_\mathcal{H}}\) where \(\mathcal{X} = \{-1,1\}\). For \ak{specified} parameters \(\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}\), \(\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N\), and \(\boldsymbol \theta_N^{\mathcal{HV}}\) as a real matrix with dimension \(N_\mathcal{H} \times N\), the RBM model for \(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) has the joint probability mass function

@@

\end{equation}
with normalizing function
\[

@@

\]
Let \(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}\), with \(q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}\), denote \ak{a given} parameter vector for the RBM, as indexed by the number \(N\) of visible random variables (which may differ from the actual lengths of these parameter vectors). The probability mass function for the visible variables \(\boldsymbol X = (X_1, \dots, X_N)\) follows from marginalizing the joint specification to yield

@@

\end{equation}
Here the baseline model \eqref{eq:RBM1} for hidden/visible variables is a linear exponential one in sufficient statistics \((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) using \(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) from \eqref{eq:RBM1}, but the form differs from the previous exponential models in \eqref{eq:expo} in that the lengths of parameters \(\boldsymbol \theta_N\) and statistics \((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) increase to incorporate more visible variables. That is, in contrast to \eqref{eq:expo}, the natural parameter function involved in the RBM model \eqref{eq:RBM1}, as the identity mapping of the parameters \(\boldsymbol \theta_N\in\mathbb{R}^{q(N)}\), naturally grows in dimension \(q(N)\to \infty\) to accommodate visible variables \(X_1, \dots, X_N\) of increasing sample size \(N\to \infty\). Additionally, one may further arbitrarily choose the number \(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RBM model \eqref{eq:RBM1} to define a marginal model \eqref{eq:RBM2} for the \(N\) visible variables \(\boldsymbol X\), and the number \(N_\mathcal{H}\) of hiddens may also potentially increase with \(N\). Because \(|\mathcal{X}| = 2\) and \(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all \(\boldsymbol x \in \mathcal{X}^N\), the RBM specification \eqref{eq:RBM2} for visibles \(\boldsymbol\) corresponds to a FOES model, while the joint distribution \eqref{eq:RBM1} for \((\boldsymbol X, \boldsymbol H)\) is also a FOES model. As this example also indicates, any model formed by marginalizing a base FOES model class, such as the RBM joint specification \eqref{eq:RBM1}, is again a FOES model.

@@

Consider two models with ``deep architecture'' that contain multiple hidden (or latent) layers in addition to a visible layer of data, namely a deep Boltzmann machine \citep{salakhutdinov2009deep} and a deep belief network \citep{hinton2006fast}. Let \(M\) denote the number of hidden layers included in the model and let \(N_{(H,1)}, \dots, N_{(H,M)}\) denote the numbers of hidden variables within each hidden layer. Then the random vector \(\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}\) collects both the hidden variables \(\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}\) and visible variables \(\boldsymbol X =(X_1,\ldots,X_N)\) in a deep probabilistic model. Each variable outcome will again lie in \(\mathcal{X} = \{-1,1\}\).

@@

\textbf{Deep belief network (DBN).} A DBN resembles a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers. The difference between the DBM and DBN models is that all but the last stacked layer \ak{(layer $0$, which contains the visible random variables $X_1, \dots, X_N$)} in a DBN are Bayesian networks \citep[see][]{pearl985bayesian}, rather than RBMs. A Bayesian network is a probabilistic graphical model that defines conditional dependence to be directed, rather than undirected (as with the RBM). Thus for visibles \(X_1, \dots, X_N\) with support \(\mathcal{X}^N, \mid \mathcal{X} \mid < \infty\), a DBN is also a FOES model with \(q(N)\) the length of parameter vector is dependent on the dimension of the visibles because \(P_{\boldsymbol \theta_N}(\boldsymbol x)>0\) for all \(\boldsymbol x \in\mathcal{X}^N\). Commonly, as in logistic belief nets \citep{neal1992connectionist}, a ``weight'' parameter is placed on each interaction between visibles, \(X_1, \dots, X_N\), and the first layer of latent variables, \(H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}\), satisfying the definition of a FOES model.

@@

To define a measure of instability in FOES models, it is useful to consider the behavior of data models \(P_{\theta_N}\), again supported on a set \(\mathcal{X}^N\) of outcomes for \(\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)\) \textbackslash ak\{and for a prescribed configuration of parameters \(\boldsymbol \theta_N\), in connection to the sample size \(N\). A relevant quantity to this end is a log-ratio of extremal probabilities (LREP), defined as
\begin{align}
\label{eq:elpr}
 \REP (\boldsymbol \theta_N)  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right],
\end{align}
based on maximum and minimal model probabilities. In what follows, the main idea is that instability, and other negative model features, can be associated with a FOES model formulation for \(N\) random variables where the LREP \eqref{eq:elpr} is overly large relative to the sample size \(N\). That is, a sequence of FOES probability models \(P_{\boldsymbol \theta_N}\) results in specifying \ak{a} distribution of observations \(\boldsymbol X=(X_1,\ldots,X_N)\) for each sample size \(N \geq 1\) \ak{(dependent on the prescribed values for $\boldsymbol \theta_N$, $N \geq 1$)} and instability will generally occur among these models whenever the corresponding LREP \eqref{eq:elpr} grows faster than \(N\). This leads to the following definition.

\BeginKnitrBlock{definition}[S-unstable FOES model]
\protect\hypertarget{def:instabFSFS}{}{\label{def:instabFSFS} \iffalse (S-unstable FOES model) \fi{} }A FOES model formulation \ak{$P_{\boldsymbol \theta_N}$} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\), \ak{based on parameter specification $\boldsymbol \theta_N$, $N \geq 1$,} is \emph{Schweinberger-unstable} or \emph{S-unstable} if
\begin{equation}
\label{eq:Sun}
\lim \limits_{N \rightarrow \infty} \frac{1}{N} \REP(\boldsymbol \theta_N) \equiv \lim \limits_{N \rightarrow \infty} \frac{1}{N}\log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right] = \infty
\end{equation}
as the number of variables increases (\(N \rightarrow \infty\)).
\EndKnitrBlock{definition}

@@

\]

@@

\end{align}
